\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-07-30 docs/parco13/paper.tex
More on the paper

\newcommand{\Fig}[1]{Figure~\ref{fig:#1}}

@@

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a uniform mesh.
  Top: run time of time integration.  Bottom: time spent in local and parallel
  neighbor exchanges.  The number of patches is $2^{2 \ell}$ at level $\ell$.
  The flat lines for the smaller runs are caused by having
  less patches than MPI processes, leaving some of them idle.  For each data point
  we use the maximum time required across all processes.}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniform-mesh experiments run on TACC's
\emph{Stampede} supercomputer at different levels of resolution.  First of all,
we use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see Figure~\ref{fig:uniscale}.  Due to space constraints we omit the
corresponding weak scaling results, which are nearly optimal as well.

The results in this section are preliminary and will be extended in the final
version.  We have so far exercised \forestclaw for two 2D manifolds,
% namely the
% unit square and a sphere composed of two mapped squares.
% On both we solve the
% advection equation with a given velocity field.
advecting a concentration field with a sharp gradient that we track dynamically
with adaptive mesh refinement.  We observe that the refinement follows the
front and allows for an accurate representation of its location over time.
% \Fig{results2d}.

\pforest is parallelized with MPI and well tested on large supercomputers, both
in 2D and 3D.  Parallelization of \forestclaw is not completed yet but
envisioned for the final version of the paper.

>>>>

\newcommand{\Fig}[1]{Figure~\ref{fig:#1}}
\newcommand{\Tab}[1]{Table~\ref{tab:#1}}

@@

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a uniform mesh.
  Top: run time of time integration.  Bottom: time spent in local and parallel
  neighbor exchanges.  The number of patches is $2^{2 \ell}$ at level $\ell$.
  The flat lines for the smaller runs are caused by having
  less patches than MPI processes, leaving some of them idle.  For each data point
  we use the maximum time required across all processes.}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniform-mesh experiments run on TACC's
\emph{Stampede} supercomputer at different levels of resolution.  First of all,
we use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.  Due to space constraints we omit the
corresponding weak scaling results, which are nearly optimal as well.

\begin{table}
  \begin{center}
    \begin{tabular}{cccc|rr}
      \multicolumn{4}{c|}{strategy} & \multicolumn{2}{c}{wall clock time} \\
      mesh & remesh & partition & time step & $P = 16$ & $P = 256$ \\
      \hline
      uniform & none       & by count  & global   & 3961.  & 256.  \\
      \hline
          AMR & every step & by count  & global   &  252.  &  54.6 \\
          AMR & every 4    & by count  & global   &  178.  &  39.7 \\
      \hline
          AMR & every step & by count  & subcycle &   99.9 &  17.3 \\
          AMR & every 4    & by count  & subcycle &   87.2 &  14.0 \\
      \hline
          AMR & every step & by weight & subcycle &   95.7 &  18.2 \\
          AMR & every 4    & by weight & subcycle &   84.4 & 14.2 \\
      \hline
    \end{tabular}
  \end{center}
\caption{Comparison of different meshing strategies for the advection example
  with a fixed maximum level 8, run on 16 and 256 cores, respectively.  The
  wall clock time includes the whole run of the program beginning to end (file
  I/O disabled).  The adaptive runs converge to a minimum level of 3.}
\label{tab:swirlwall}
\end{table}
Next, we examine how we can reduce the overall wall clock time of a simulation
by switching from uniform refinement to AMR with the same maximum level,
effectively coarsening where high resolution is not beneficial.  For an example
with a sharp front as depicted in \Fig{results2d} we expect considerable
savings by AMR, which we confirm in \Tab{swirlwall}.  The wall clock times can
be reduced by up to a factor of 50 by combining AMR with subcycling.  The
adaptive runs are still very well load balanced overall, which we infer from
the linear scaling of the time integration alone (not shown) and from the fact
that the weighted partition does not influence the run times.  There is the
tradeoff however, at least in the current implementation, that the adaptive
runs suffer from the loss in scalability associated with the neighbor exchanges
that we discussed earlier.

% The results in this section are preliminary and will be extended in the final
% version.  We have so far exercised \forestclaw for two 2D manifolds,
% % namely the
% % unit square and a sphere composed of two mapped squares.
% % On both we solve the
% % advection equation with a given velocity field.
% advecting a concentration field with a sharp gradient that we track dynamically
% with adaptive mesh refinement.  We observe that the refinement follows the
% front and allows for an accurate representation of its location over time.
% % \Fig{results2d}.
% 
% \pforest is parallelized with MPI and well tested on large supercomputers, both
% in 2D and 3D.  Parallelization of \forestclaw is not completed yet but
% envisioned for the final version of the paper.
